Deep Reinforcement Learning for Conservation Decisions

### 2021-05-12 manuscript/manuscript.Rmd
Getting yesterday's additions up

<!-- Environmental problems, ML -->
Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@worm2006], declining biodiversity [@], and zoonotic pandemics [@covid].

<!-- citation suggestion for declining biodiversity  -->

<!-- @article{dirzo2014defaunation, -->
<!--   title={Defaunation in the Anthropocene}, -->
<!--   author={Dirzo, Rodolfo and Young, Hillary S and Galetti, Mauro and Ceballos, Gerardo and Isaac, Nick JB and Collen, Ben}, -->
<!--   journal={science}, -->
<!--   volume={345}, -->
<!--   number={6195}, -->
<!--   pages={401--406}, -->
<!--   year={2014}, -->
<!--   publisher={American Association for the Advancement of Science} -->
<!-- } -->

But will ML approaches really help us tackle a changing planet?

<!-- suggestion: "...tackle our most pressing environmental challenges" -->

Applications of ML in ecology have begun to illustrate the promise of two methods of ML: *supervised learning* [@joseph2020] and *unsupervised learning* [@unsupervised], but have so far largely overlooked the third and possibly most promising approach in the ML triad: *reinforcement learning* (RL).
Three features distinguish RL from other ML methods in ways that are particularly well suited to addressing global ecological change issues:

@@

<!-- really like this- super clear and helpful! -->

Despite relevance to uncertainty and decision making that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in this area.
To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@lillicrap; @alphazero; @alphastar]
Complex environmental problems share many similarities to these tasks and games: the need to plan many moves ahead given a large number of possible outcomes, to account for uncertainty and to respond with contingency to the unexpected.
RL agents typically develop their strategies by interacting with simulators, a practice that should not be unsettling to ecologists since learning from simulators is common across ecology.
Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or global circulation models (GCMs) in climate [@gcm] are already used to explore scenarios and inform ecosystem management.

<!-- another widely used model is ED2 (ecosystem demography model) - I think more common than SORITE? https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2008JG000812 -->

Even simpler simulators are frequently employed throughout ecology to guide understanding and response.
<!-- Kari: ick, jargon heavy.  Need to say: some existing methods, but limited to simple problems or simple solutions -->
Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].

<!-- I think there is a missing the connection between the examples and needing more sophisticated methods for decision making below. i.e. "As we develop more complex process based models (such as Ecosim and ED2), methods that allow us to effectively integrate these high dimensional models into decision making processes will be critical to ensuring their policy relevance." (or something like that, then go on to NN ML)-->

Neural-network-based RL techniques, referred to as *deep RL*, have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
We believe that the application of RL methods to simulations of ecological processes will open the door to a new paradigm for tackling tough environmental change problems.
We also believe that such a transformation could bring risks and pitfalls comparable and varied as its solutions.

@@

 <!-- I like the summary but could maybe be less specific> "We start with a simple example from fisheries management that has a known optimal solution, proving that with no prior knowledge of underlying model dynamics RL agents can find a close to optimal solution. We go on to show that RL can outperform rule-of-thumb solutions to management scenarios not amenable to classic optimization methods such as dynamic programming. Our examples demonstrate that..."  -->

@@

<!-- maybe cite Caleb's paper here? -->

<!-- these last few sentences almost feel like discussion points. I'm torn though, because it it nice to have them in the intro. Maybe condense into 1 sentence at the end of you "summary" paragraph of the bulk of the paper above. for example: "We also explore the potential pitfalls of RL-based solutions to global environmental change problems. We show how, and why, these challenges are not only technical, but include issues of ethics and power, especially when the algorithms or data are proprietary." then add a longer discussion towards the end of the paper-->

@@

<!-- something to introduce this section/transition from the intro might be helpful: "We start with a brief overview of RL problems and terminology:" -->

@@

<!-- ^ this is great! I do think keeping a table of these terms would be helpful (and refering to that table in the text here) -->

@@

Amongst model-free algorithms, there is divergence around what functions the agent is attempting to learn.
Generally, model-free agents are either learning a policy function, a value function or both.
For context, *value functions* are proxies for how high of a cumulative reward an agent can expect to receive from a given state or state-action pair.
If an agent knows the value function, then the agent can find the optimal action at any state by selecting the action that will maximize the value function in expectation.
The class of algorithms that exclusively try to learn a value function are called *value learning* methods.
By contrast, *Policy gradient* algorithms exclusively learn the policy.
*Actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth -- for more, see [@CITE]; yet, the main point of distinction comes from the bias-variance trade-off.
Policy gradients tend to struggle with high variance, value-learning tends to be biased while actor-critic methods balance these two concerns.

@@

An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of it's current *state* and associated *reward* as output.
In contrast to classical approaches, such as dynamic programming, there are few restrictions on what comprises a state or action, at least in principle.
The state may be continuous or discrete, completely or only partially observed, one or many dimensions.
For instance, we might represent a ecological community of 10 interacting species using an environment with 10 continuously valued states representing the possible population densities of each species.
The action space may similarly complex, and the response of the environment to any possible action may be deterministic or stochastic in nature.
The reward will in general depend on both the current state and choice of actions.
An environment must also specify a stopping condition, such as when a particular state is reached or a pre-defined interval of time has elapsed.
In some environments, a reward is only received after the agent successfully completes a potentially long sequence of complex actions, such as solving a maze.
In other environments, an agent may have to learn to forgo immediate reward (or even accept an initial negative reward or cost) in order to maximize the net discounted reward over the simulation period, as we illustrate in examples here.

@@

<!-- 
Placeholder table as figure, rather hacky (pandoc doesn't like `longtable`) 
Make a nice table box in latex instead if we're keeping this
-->

<!-- maybe consider some combined conceptual figure/table?  But overall I really like the table and think it will be helpful for readers -->

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{table1.pdf}
\end{figure}

@@

In formal terms, the goal of reinforcement learning is to find an optimal policy distribution, $\pi^*(a_t | s_t)$ or $\pi^*(a_t|o_t)$ for MDP and POMDP cases respectively, that maximizes the expected cumulative, discounted reward:

$$
  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big]
$$

@@

Many reinforcement learning algorithms center around estimating value functions, which attempt to quantify the goodness of being in a state or taking an action from a state.
Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to the likelihood that a state or state-action pair will result in a high cumulative reward trajectory.
There are 3 value functions that are commonly used across RL algorithms: the state-value function, $V^\pi(s_t)$, the state-action value function, $Q^\pi(s_t, a_t)$, and the advantage function, $A^\pi(s_t, a_t)$.
The state-value function is defined as the expected return from a given state,

@@

<!-- second pass at this -->

Algorithms are playing increasingly influential decision-making roles in sectors such as policing, finance, and employment, raising concerns about the ethics, biases, and power entrenched in both the algorithms and their outputs [@kalluri2020]. As algorithms play a greater role in environmental and conservation policy, will similar concerns arise? While this question has garnered  recent interest in the literature [@wearn2019; @adams2019; @scoville2021] the use use of decision making algorithms in environmental applications has largely been confined to the context of classification algorithms (i.e. to identify wildlife poaching [@adams2019] or illegal fishing behavior [@mcdonald2021]). Increasingly, applications of AI to meeting sustainable development goals [@], prioritizing protected area locations [@], and predicting However, good intentions are not sufficient to ensure equitable outcomes. Interdisciplinary collaborations with experts in algorithmic governance and ethics will be important ensure that applications of AI to ecological challenges are applied responsibly.

>>>>

<!-- Environmental problems, ML -->
Advances in both available data and computing power have begun to open the door to a greater role for machine learning (ML) in addressing some of our planet’s most pressing environmental problems, such as the growing frequency and intensity of wildfire [@wildfire], over-exploited fisheries [@worm2006], declining biodiversity [@], and zoonotic pandemics [@covid].
But will ML approaches really help us tackle a changing planet?
Applications of ML in ecology have begun to illustrate the promise of two methods of ML: *supervised learning* [@joseph2020] and *unsupervised learning* [@unsupervised], but have so far largely overlooked the third and possibly most promising approach in the ML triad: *reinforcement learning* (RL).
Three features distinguish RL from other ML methods in ways that are particularly well suited to addressing global ecological change issues:

@@

Despite relevance to uncertainty and decision making that could make RL uniquely well suited for ecological and conservation problems, it has so far seen little application in this area.
To date, the problems considered by RL research have largely been drawn from examples in robotic movement and games like Go and Starcraft [@lillicrap; @alphazero; @alphastar]
Complex environmental problems share many similarities to these tasks and games: the need to plan many moves ahead given a large number of possible outcomes, to account for uncertainty and to respond with contingency to the unexpected.
RL agents typically develop their strategies by interacting with simulators, a practice that should not be unsettling to ecologists since learning from simulators is common across ecology.
Rich, processes-based simulations such as the SORTIE model in forest management [@sortie], Ecopath with Ecosim in fisheries management [@ecopath], or global circulation models (GCMs) in climate [@gcm] are already used to explore scenarios and inform ecosystem management. 
Even simpler simulators are frequently employed throughout ecology to guide understanding and response.
<!-- Kari: ick, jargon heavy.  Need to say: some existing methods, but limited to simple problems or simple solutions -->
Decision-theoretic approaches based on optimal control techniques can only find the best strategy in the simplest of ecological models; the so called "curse of dimensionality" makes problems with a large number of states or actions intractable by conventional methods such as stochastic dynamic programming (SDP) and approximate dynamic programming (ADP) [@Polasky2011; @Marescot2013].
Neural-network-based RL techniques, referred to as *deep RL*, have proven particularly effective in problems involving complex, high-dimensional spaces that have previously proven intractable to classical methods.
We believe that the application of RL methods to simulations of ecological processes will open the door to a new paradigm for tackling tough environmental change problems.
We also believe that such a transformation could bring risks and pitfalls comparable and varied as its solutions.

@@

@@

@@

@@

@@

Amongst model-free algorithms, there is divergence around what functions the agent is attempting to learn.
Generally, model-free agents are either learning a policy function, a value function or both.
For context, *value functions* are proxies for how high of a cumulative reward an agent can expect to receive from a given state or state-action pair.
If an agent knows the value function, then the agent can find the optimal action at any state by selecting the action that will maximize the value function in expectation.
The class of algorithms that exclusively try to learn a value function are called *value learning* methods.
*Policy gradient* algorithms exclusively learn the policy.
*Actor-critic* algorithms attempt to learn both a policy function and a value function, whereby the value function, which is learned separately from the policy, is used to inform the agent on the goodness of a selected action.
There is a handful of advantages and disadvantages for each of these approaches that we will not go in depth -- for more, see [@CITE]; yet, the main point of distinction comes from the bias-variance trade-off.
Policy gradients tend to struggle with high variance, value-learning tends to be biased while actor-critic methods balance these two concerns.

@@

An environment is a mathematical function, computer program, or real world experience that takes an agent's proposed *action* as input and returns an observation of it's current *state* and associated *reward* as output.
In contrast to classical approaches, there are few restrictions on what comprises a state or action, at least in principle.
The state may be continuous or discrete, completely or only partially observed, one or many dimensions.
For instance, we might represent a ecological community of 10 interacting species using an environment with 10 continuously valued states representing the possible population densities of each species.
The action space may similarly complex, and the response of the environment to any possible action may be deterministic or stochastic in nature.
The reward will in general depend on both the current state and choice of actions.
An environment must also specify a stopping condition, such as when a particular state is reached or a pre-defined interval of time has elapsed.
In some environments, a reward is only received after the agent successfully completes a potentially long sequence of complex actions, such as solving a maze.
In other environments, an agent may have to learn to forgo immediate reward (or even accept an initial negative reward or cost) in order to maximize the net discounted reward over the simulation period, as we illustrate in examples here.

@@

<!-- 
Placeholder table as figure, rather hacky (pandoc doesn't like `longtable`) 
Make a nice table box in latex instead if we're keeping this
-->
\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{table1.pdf}
\end{figure}

@@

In formal terms, the goal of reinforcement learning is to find an optimal policy distribution, $\pi^*(a_t | s_t)$ or $\pi^*(a_t|o_t)$ for MDP and POMDP cases respectively, that maximizes the expected return, $J(\pi)$:

$$
  \pi^* = \underset{\pi}{\text{argmax}} \,\, \mathbb{E}_{\tau \sim p_\pi(\tau)}\Big[\sum_{t=0}^{H} \gamma^t r(s_t, a_t) \Big] =  \underset{\pi}{\text{argmax}} \,\, J (\pi)
$$

@@

<!-- Explain policy gradients as this is the natural progression
So sketch out reinforce, ending that its susceptibility to high variance

talk about value-based learning so q values and bellman recursion
-->
## Policy Gradients
The most straightforward way to optimize the RL objective is through a policy search, whereby the agent continually updates a policy parameterized by $\theta$, $\pi_\theta$ -- in deep RL, this parameterized policy will be a neural network.
There are both gradient-free and gradient-based policy search methods that have been successful in practice [@CITE; @CITE]; but for the sake of brevity, we will focus on gradient-based methods, which are more commonly encountered. 
In policy gradient methods, algorithms are performing gradient ascent on the expected return, $J(\pi_\theta)$, to find the optimal policy parameters. 
At each gradient step, the following update is performed,

$$
  \theta_{t+1} = \theta_t + \alpha \nabla_\theta J(\pi_\theta).
$$

Supposing that the agent has interacted with the environment and has collected some trajectories denoted by, $\tau_i$, then $\nabla J(\pi_\theta)$ can be estimated as:

$$
  \nabla_\theta J(\pi_\theta) \approx \sum_{i,t} \nabla_\theta \log{\pi_\theta (a_{i,t} \mid s_{i,t})} \Big( \sum_{t'=t}^H \gamma^{t'-t} r(s_{i, t'}, a_{i, t'}) \Big).
$$

See the appendix for how this is derived. 
The intuition behind this approximation is exactly what we want: the agent will increase the probability of repeating the actions for the inputted states proportional to the size of the return. 
In practice, however, the empirical return tends to have high variance, resulting in a very noisy gradient ascent that impedes learning [@CITE].
To avoid this issue, most policy gradient algorithms replace the empirical return with a different return estimator that has lower variance [@CITE GAE].
A sketch of a simple policy gradient algorithm is shown in FIGURE.

## Value Learning
Another way to approach the RL problem is by estimating a value function, and using the value function to retrieve a policy. 
Value functions attempt to quantify the goodness of being in a state or taking an action from a state.
Given that the objective of RL is to maximize the cumulative expected reward, "goodness" refers to the likelihood that a state or state-action pair will result in a high cumulative reward trajectory.
The state-action-value or Q function finds the expected return provided a state and action,

$$
  Q^\pi(s_t, a_t) = \mathbb{E}_{\tau \sim p_\pi(\tau \mid s_t, a_t)}\Bigg[\sum_{t' = t}^H \gamma^{t' - t} r(s_t, a_t)\Bigg].
$$

The general motivation is that if we know the Q function, then we can easily find the optimal action at any state by selecting the action with the highest Q-value: $a_t^* = \underset{a_t}{\text{argmax}} \, Q^\pi (s_t, a_t)$.
In the case that the state-action space can be tractably represented in a table, one could employ an algorithm, like temporal-difference control -- see Sutton Ch. 6 for details -- which approximates the Q value for every state-action pair [@suttonbarto]. 
Yet, for many decision-making problems, the state-action space cannot be tractably represented in a table; one effective alternative for this case is to use a neural network to approximate the Q-function.

There are 3 value functions that are commonly used across RL algorithms: the state-value function, $V^\pi(s_t)$, the state-action value function, $Q^\pi(s_t, a_t)$, and the advantage function, $A^\pi(s_t, a_t)$.
The state-value function is defined as the expected return from a given state,

@@
